Tropical Mathematics for the Lambda Calculus

### 2023-01-19 appIII.tex
commit

In the following we let $Q$ indicate the Lawvere quantale $([0,\infty], \geq, +,0)$. 
$Q$ can be seen as:
\begin{itemize}
\item a category $Q_{0}$ with a unique object and morphisms being the elements of $Q$, with identity 0 and composition $+$;

\item a degenerate category $Q_{1}$ with objects the elements of $Q$, and  
a morphism between $x$ and $y$ whenever $x\geq y$;

\item a 2-category $Q_{2}$ with 0-cells and 1-cells as in $Q_{0}$, and 2-cells the morphisms of $Q_{1}$.

\end{itemize}
These categorical formulations of $Q$ will help us here and there.

We let $x-y$ indicate truncated subtraction. Notice that 
$x \leq (x-y)+y$ and $(x+y)-y\leq x$. This corresponds to the fact that ``$\_-y$'', as a functor from $
Q_{1}$ to itself, 
 is \emph{right-adjoint} to the functor ``$\_+y$''. Equivalently, 
 $x+y \geq z $ iff $x\geq z-y$.

$Q\mathsf{Rel}$ is the category of sets and $Q$-relations, where a $Q$-relation 
$R:X\pfun Y$ is a function $R: X\times Y\to Q$, with identity $1_{X}:X\pfun X$ given by 
$$1_{X}(x,y)=\begin{cases} 0 & \text{ if }x=y\\ \infty & \text{ otherwise}\end{cases}$$
and composition of $R:X\pfun Y$ and $S:Y\pfun Z$ given by 
$$
(S\cdot R)(x,z)= \inf_{y\in Y}R(x,y)+S(y,z)
$$

\begin{remark}
$Q\mathsf{Rel}$ is also a (degenerate) 2-category, where for $R,S:X\pfun Y$, $R\To S$ iff for all $x\in X$, $y\in Y$, $R(x,y)\geq S(x,y)$.

\end{remark}

Composition with $R:X\pfun Y$ in Q$\mathsf{Rel}$ yields a functor
$$
R\cdot \_ : Q\mathsf{Rel}(A,X)   \longrightarrow Q\mathsf{Rel}(A,Y)
$$
which
has a right-adjoint
$$
\_ \multimapinv R : Q\mathsf{Rel}(A,Y)   \longrightarrow Q\mathsf{Rel}(A,X)
$$
given, for $S:A\pfun Y$ by 
\begin{align*}
(S\multimapinv R)(a,x)= \sup_{y\in Y}S(a,y) -R(x,y)
%(R\multimap S)(a,x)= \sup_{y\in Y}
\end{align*}

%
%\begin{lemma}
%\begin{itemize}
%\item[i.] $ (R\multimapinv T)\cdot
%(T\multimapinv S)\leq R \multimapinv S $.
%
%\item[ii.] $R\multimapinv T \leq (R\multimapinv S)\multimapinv (T\multimapinv S)$.
%\end{itemize}
%\end{lemma}
%\begin{proof}
%\begin{align*}
%(R\multimapinv S)(x,y)&= \sup_{z}R(x,z)-S(y,z) \\
%&=\sup_{z} (R(x,z)-T(z,z))+(T(z,z)-S(y,z))\\
%&= \inf_{w}\left(\sup_{z}R(x,z)-T(w,z)\right) +\left(\sup_{z'}T(w,z')-S(z',y)\right)
%\\
%&= \inf_{w}(R\multimapinv T)(x,w)+(T\multimapinv S)(w,y)  
%\\
%&= (R\multimapinv T)\cdot (T\multimapinv S)(x,y)
%\end{align*}
%
%\end{proof}

\subsubsection{$Q$-categories.}

A \emph{$Q$-category}, or \emph{generalized (quasi-)metric space} is a small $Q$-enriched category. Concretely, it is given by a pair $(X,q)$ made of a set $X$ and a $Q$-relation $q: X\pfun X$ satisfying:
\begin{align}
q(x,x) & = 0 \tag{Q1} \\
q(x,z)+q(z,y) & \geq q(x,y) \tag{Q2}
\end{align}
A $Q$-category is \emph{skeletal} if it further satisfies
\begin{align}
q(x,y) & = 0 \ \To \ x=y \tag{Q3}
 \end{align}
 and \emph{symmetric} if it satisfies
 \begin{align}
q(x,y) & = q(y,x) \tag{Q4}
 \end{align}
For simplicity, in the following we indicate the distance function of a $Q$-category $X$ simply as $X(x,y)$.

\begin{remark}
In more abstract terms, a $Q$-category can be identified with a \emph{monad} in $Q\mathsf{Rel}$, seen as a 2-category: it is given by a 1-cell $q:X\pfun X$ together with 2-cells $1_{X}\To  q$ and $q\cdot q \To q$. 
\end{remark}

We let $x\simeq y$ indicate the equivalence relation defined by  $X(x,y)=X(x',y)$ holds for all $y\in X$. 
Notice that $x\simeq y$ coincides with $x=y$ precisely when $X$ is skeletal.

Every $Q$-category $X$ is also a preorder $(X,\preceq_{X})$, with $x\preceq_{X}y$ iff $X(x,y)=0$. Moreover, if $X$ is skeletal, then $\preceq_{X}$ is an order.

For every $Q$-category $X$, the category $X^{\mathrm{op}}$ has same objects as $X$, and $X^{\mathrm{op}}(x,y)=X(y,x)$.

\begin{example}
$Q$ is a $Q$-category with the distance $Q(x,y)=y-x$. 
The category $Q^{\mathrm{op}}$ has distance $Q^{\mathrm{op}}(x,y)=Q(y,x)$.
\end{example}

\begin{example}
For every $Q$-category $X$, the distance $\C S(X)(x,y)=\max\{X(x,y),X^{\mathrm{op}}(x,y)\}$ defines a symmetric $Q$-category $\C S(X)$. In the case of $Q$, $\C S(Q)$ is just the Euclidean distance.
\end{example}

A \emph{functor} between $Q$-categories $X$ and $Y$ is just a non-expansive function $f:X\to Y$, that is, satisfying $Y(f(x),f(y))\leq X(x,y)$.

%
\begin{remark}\label{rem:functor}
Given $f:X\to Q$, $f$ is a functor precisely when for all $x,y\in X$ $f(x)\leq\inf_{x'\in X}f(x')+X(x',x)$. 
Indeed, if $f$ is a functor then $f(x)\leq f(x')+X(x',x)$, since $f(x)-f(x')=Q(f(x'),f(x))\leq X(x',x)$. 
Conversely, if $f(x)\leq f(x')+X(x',x)$ holds for all $x'$, then 
$Q(f(x),f(x'))=f(x')-f(x)\leq X(x',x)$. 
\end{remark}

%Indeed, being a functor means that any arrow $X(x,y)$ is mapped into an arrow $Y(f(x),f(y))$, that is, $Y(f(x),f(y))

\begin{example}
For every $Q$-category $X$, the $Q$-category $[X,Q]$ has objects all functors from $X$ to $Q$ (where the latter is seen as a $Q$-category), and $[X,Q](f,g)=\sup_{x\in X}Q(f(x),g(x))$.

When $X$ has the discrete metric, $[X,Q]=Q^{X}$ and $\C S([X,Q])(f,g)=\| f-g\|_{\infty}$.

\end{example}

\begin{example}
For every $Q$-categories $X,Y$ we can define:
\begin{itemize}
\item a $Q$-category $X\otimes Y$ with 
$(X\otimes Y)(\langle x,y\rangle, \langle x',y'\rangle)=X(x,x')+Y(y,y')$;
\item a $Q$-category $X\times Y$ with 
$(X\times Y)(\langle x,y\rangle, \langle x',y'\rangle)=\max\{X(x,x'),Y(y,y')\}$. 
\end{itemize}
\end{example}

\subsubsection{$Q$-distributors.}

%We let $Q\mathsf{Cont}$ indicate the category of $Q$-categories and continuous functors.

A \emph{$Q$-distributor} $\Phi: X\pfun Y$ is a $Q$-relation $\Phi: X\times Y\to Q$ satisfying
\begin{align*}
\Phi\cdot X   \leq\Phi \qquad \qquad
Y \cdot \Phi  \leq \Phi
\end{align*}

\begin{remark}
Using Remark \ref{rem:functor}, it follows that $\Phi$ is a $Q$-distributor precisely when it is a functor $\Phi: X^{\mathrm{op}}\otimes Y \to Q$:
the condition $\Phi(x,y) \leq X(x,x')+\Phi(x',y)$ implies that $\Phi$ is a functor over $X^{\mathrm{op}}$, while
the condition $\Phi(x,y) \leq \Phi(x,x')+Y(x',y)$ implies that $\Phi$ is a functor over $Y$.
But this implies then 
$\Phi(x,y) \leq X(x,x') +\Phi(x',y')+Y(y',y)$ and thus $Q(\Phi(x',y'),\Phi(x,y))\leq X(x,x')+Y(y',y)$.
\end{remark}

%where $(\Phi\cdot X)_{x,y}= \inf_{z\in X}\Phi_{x,z}+X(z,y)$ and 
%$(y\cdot \Phi)_{x,y}= \inf_{z\in X}Y(x,z)+\Phi_{z,y}$.

The identity distributor $1_{X}:X\pfun X$ is given by $(1_{X})_{x,y}=X(x,y)$. 
Distributors $\Phi:X\pfun Y$ and $\Psi:Y\pfun Z$ compose via $\Psi\cdot \Phi:X\pfun Z$. 
We let $Q\mathsf{Dist}$ indicate the category of $Q$-categories and $Q$-distributors.

\begin{remark}
Usually distributors $\Phi: X\pfun Y$ are presented as $Q$-relations $\Phi:Y\times X\to Q$ (notice the inversion of $X$ and $Y$), that is, as functors $\Phi: X\times Y^{\mathrm{op}}\to Q$. We chose to invert the presentation of distributors for uniformity with the usual notations in models of the differential $\lambda$-calculus. 
\end{remark}

Any functor $f:X\to Y$ induces two adjoint distributors $f^{\circ}:X\pfun Y$ and $f_{\circ}:Y\pfun X$ given by $f^{\circ}(x,y)=Y(f(x),y)$ and $f_{\circ}(y,x)=Y(f(y),x)$.
Here adjoint means that $1_{X}\leq f_{\circ}\cdot f^{\circ} $ and $f^{\circ}\cdot f_{\circ}\geq 1_{Y}$.

\begin{remark}[Yoneda embedding]
%$Q$ is a $Q$-category with $Q(x,y)=|x-y|$. 
%For any $Q$-category $X$ we can define a new $Q$-category $\mathsf{Dist}(X,Q)$ (that we note simply as $[X,Q]$) whose objects are the distributors $\B x:\B 1\pfun X$, or equivalently, the functors from $X$ to $Q$, i.e.~those $\B x\in Q^{X}$ such that $|\B x_{a}-\B x_{b}|\leq X(a,b)$, or equivalently $\B x_{a}+X(a,b) \geq \B x_{b}$.

The \emph{Yoneda embedding} is the faithful functor $\C Y: X\to [X,Q]$ given by $\C Y(x)(y)=X(y,x)$. The functoriality and faithfulness of $\C Y$ follow from
\begin{align}
[X,Q](\C Y(x),\C Y(x'))= X(x,x') \tag{Yoneda}
\end{align}
which is proved as follows: for all $y\in X$ we have 
\begin{align*}
Q( \C Y(x)(y),\C Y(x')(y))&= \C Y(x')(y)-\C Y(x)(y) \\
&= X(y,x')-X(y,x) \leq X(x,x')
\end{align*}
where the last step follows from $X(y,x)+X(x,x')\geq X(y,x')$. 
From this we deduce that $[X,Q](\C Y(x),\C Y(x'))=\sup_{y\in X}Q( \C Y(x)(y),\C Y(x')(y))\leq X(x,x')$. 
For the converse direction, we have  
\begin{align*}
X(x,x') &  = X(x,x') - 0 \\ &=
X(x,x')- X(x,x)
\\& =\C Y(x')(x)-\C Y(x)(x)= Q(\C Y(x)(x),\C Y(x')(x))\\
&\leq [X,Q](\C Y(x), \C Y(x'))
\end{align*}
\end{remark}

\begin{remark}
The \emph{opposite Yoneda embedding} is the faithful functor
$\C Y^{\mathrm{op}}: X\to [X,Q]^{\mathrm{op}}$ given by $\C Y^{\mathrm{op}}(x)(y)=X(x,y)$. The functoriality and faithfulness of $\C Y^{\mathrm{op}}$ follow from
\begin{align}\label{eq:yonedaop}
[X,Q](\C Y^{\mathrm{op}}(x),\C Y^{\mathrm{op}}(x'))= X(x',x) \tag{Yoneda$^{\mathrm{op}}$}
\end{align}
which is proved similarly to the case of $\C Y$.
\end{remark}

\subsubsection{cocompleteness}

Let $\Phi:Y\pfun Z$ be a distributor and $f:Y\to X$ be a functor.
A functor $ g:Z\to X$ is called the \emph{$\Phi$-weighted colimit of $f$} if it satisfies, for all
$z\in Z$ and 
 $x\in X$, 
$$
X(g(z), x) \ = \  \sup_{y\in Y}X(f(y),x) - \Phi(y,z)
$$
If this colimit exists, we write it as $\mathrm{colim}(\Phi,f)$.

A $Q$-category $X$ is said \emph{cocomplete} if it admits all weighted colimits, and a functor $f:X\to Y$ of $Q$-categories is said \emph{cocontinuous} if it commutes with all weighted colimits, meaning that $f \circ \mathrm{colim}(\Phi,g)= \mathrm{colim}(\Phi, f\circ g)$.

We let $Q\mathsf{CCat}$ indicate the category of skeletal and cocomplete $Q$-categories and cocontinuous functors.

\begin{proposition}\label{prop:yonedasup}
Let $X$ be a $Q$-category. Then $X$ is cocomplete iff the Yoneda embedding has a left-adjoint.
\end{proposition}
\begin{proof}
 For all $\B x\in [X,Q]$, let $\sup\B x$ be defined as a weighted colimit via
$$
X(\sup\B x, y)= \sup_{z\in X} X(z,y)-\B x_{z}
$$
that is, $\sup \B x= \mathrm{colim}( \B x, \mathrm{id}_{X})$, where $\B x$ is seen as a distributor $\B x:\B 1\pfun X$.

Let us check that $\sup: [X,Q]\to X$ is a functor. First, let us check the  inequality
\begin{align}\label{eq:in1}
(X\multimapinv\B y)\cdot (\B y\multimapinv\B x)\geq (X\multimapinv\B x)\end{align}
as follows:
\begin{align*}
\left((X\multimapinv \B y)\cdot (\B y\multimapinv\B x)\right)(a)&=
 \left (\sup_{b}X(b,a)-\B y_{b}\right)+\left(\sup_{b}\B y_{b}-\B x_{b}\right)\\
&\geq
\sup_{b}(X(b,a)-\B y_{b})+(\B y_{b}-\B x_{b})
\\
&=\sup_{b}(X(b,a)-\B y_{b}+\B y_{b})-\B x_{b}
\\
&=\sup_{b} X(b,a)-\B x_{a}\\
&= (X\multimapinv \B x)(a)
\end{align*}
From \eqref{eq:in1} we deduce immediately the inequality below:
\begin{align}\label{eq:in2}
\B y\multimapinv \B x \geq (X\multimapinv \B x)\multimapinv(X\multimapinv\B y)
\end{align}
and we can now compute:
\begin{align*}
[X,Q](\B x, \B y)&= \sup_{a\in X}\B y_{a}-\B x_{a}\\
 &\stackrel{\tiny\eqref{eq:in2}}{\geq}
 \sup_{a\in X} \left( \sup_{b\in X}X(b,a)-\B x_{a}\right) -\left(  \sup_{b\in X}X(b,a)-\B y_{a}   \right )\\
 &= \sup_{a\in X}X(\sup \B x, a)-X(\sup \B y, a)\\
 &= [X,Q](\C Y^{\mathrm{op}}(\sup\B y),\C Y^{\mathrm{op}}(\sup \B x))
 \tag{\ref{eq:yonedaop}}
 \\
&=X(\sup \B x, \sup \B y)
\end{align*}

Then for all $x\in X$, $\sup \C Y(x)\simeq x$. Indeed we have 
\begin{align*}
X(\sup \C Y(x), y) & = \sup_{z\in X}X(z,y) -\C Y(x)(z) \\
& = \sup_{z\in X}X(z,y)-X(z,x)\\
&= X(x,y)
\end{align*}
%where we use the fact that from $X(x,y)+X(z,x) \geq X(z,y)$ it follows $X(x,y) \geq X(z,y)-X(z,x)$ and thus $X(x,y) \geq \sup_{z}X(z,y)-X(x,z)$, and conversely, from $X(x,y)-X(x,x)=X(x,y)$ it follows
%%$X(x,y) \leq \sup_{z}X(z,y)-X(x,z)$.
%

Moreover, for all $\B x\in [X,Q]$, we have $\C Y(\sup \B x)\geq \B x$:  
\begin{align*}
\C Y(\sup \B x)(a) &= X(a,\sup \B x)\\
&= X(  \sup(\C Y(a)),\sup \B x) \\
& \geq [X,Q]( \C Y(a), \B x)\\
&\geq\B x_{a}-\C Y(a)(a) \\
&=
 \B x_{a}-X(a,a)  = \B x_{a}
\end{align*}
%where in the last step we use the fact that 
%$[X,Q](  \C Y(a),\B x)= \sup_{b}\B x_{b}-X(a,b) = \B x_{a}$.
%
%$\B y\in Q^{X}$, 
% $\| \C Y(\sup\B x)- \B y\|_{\infty}=\| \C Y(\sup\B x)- \B x\|_{\infty}$. Indeed we have 
% \begin{align*}
% \| \B x_{a}-\C Y(\sup\B x)\|_{\infty}&= \sup_{a\in X}|
%\B x_{a}-\C Y(\sup\B x)(a)|\\ 
%&=
%\sup_{a\in X}|
%\B x_{a}-X(\sup\B x,a)|\\
% &=
%\sup_{a\in X}\inf_{a'\in X}| \B x_{a}+X(a',a) -\B x_{a'}|\\
%&= \sup_{a\in X}|\B x_{a}+ X(a,a)- \B x_{a}| = 0
% \end{align*}
%
%

Conversely, if $\sup$ is well-defined and adjoint to $\C Y$, then 
given $\Phi: Y\pfun \B 1$ and $f:Y\to X$, 
we can define
$\mathrm{colim}(\Phi, f):= \sup\Psi$, where $\Psi=
 f^{\circ}\cdot \Phi: X\pfun \B 1$, since 
\begin{align*}
X(\sup \Psi, y)&=
\sup_{z\in X}X(z,y)-\Psi(z) \\
&=\sup_{z\in X}X(z,y)- \inf_{y\in Y}X(z,f(y))  +\Phi(y)\\
& = \sup_{z\in X}\sup_{y\in Y}X(z,y)-X(z,f(y))  -\Phi(y)\\
&=
\sup_{z\in X} X(f(z),y) - \Phi(z)
\end{align*}
\end{proof}

\begin{definition}[MacNeill Completion]
Let $X$ be a $Q$-category. For all $f: \B 1\pfun X$ and $g:X\pfun \B 1$, let $f \coh g$ iff 
$f  = X\multimapinv g $ and 
$g = f\multimap X$.

The \emph{MacNeill completion of $X$} is the $Q$-category $\B M(X)$ made of those 
$f:\B 1\pfun X$ such that $f\coh g$ for some $g:X\pfun \B 1$, with
$\B M(X)(f,f')=[X,Q](f,f')$. 
\end{definition}

Observe that if $f\coh g$, then $f= X\multimapinv (f\multimap X)$, i.e.:
\begin{align}
f(x)=  \sup_{y\in X}\inf_{z\in X}X(x,y)-X(y,z) +f(z)       
\tag{COH}
\end{align}

\begin{proposition}
Let $X$ be a $Q$-category. 
If $X$ is cocomplete, then $\C Y$ is an isomorphism between $X$ and $\B M(X)$. 
\end{proposition}
\begin{proof}
For all $x\in X$, one can check that $\C Y(x)\in \B M(X)$. Indeed, we can check that $\C Y(x) \coh \C Y^{\mathrm{op}}(x)$:
\begin{align*}
\C Y(x)(y)  &= X(y,x) \\
&=  \sup_{z\in X}X(y,z  )- X(x,z)\\
&= \sup_{z\in X}X(y,z  )- \C Y^{\mathrm{op}}(x)(z)\\
 &= (X\multimapinv \C Y^{\mathrm{op}}(x))(y)
\end{align*}
Since $\sup\C Y(x) \simeq x$ holds, it suffices to show that if $\B x \coh \B y$, then 
$\C Y(\sup \B x)=\B x$: 
\begin{align*}
\C Y(\sup \B x)(a) & \ \ = X(a, \sup\B x) \\
&\ \ = \sup_{b\in X}X(a,b)-X(\sup\B x, b) \\
&\ \ = \sup_{b\in X}\inf_{c\in X}X(a,b)- X(b,c) + \B x_{c} \\
 &\stackrel{{\tiny\text{(COH)}}}{=} \B x_{a}
\end{align*}
\end{proof}

%
%\begin{remark}[other notions of cocompleteness]
%The categorical notion of cocompleteness subsumes several other notions of cocompleteness, in the sense of being (strictly) stronger:
%\begin{itemize}
%\item \emph{order-completeness} is the case when the pre-order $\preceq_{X}$ is cocomplete.   
%Let $I\subseteq X$, so that $\mathrm{id}_{I}$ can be seen as a functor from the subcategory $I$ to $X$. Let $0_{I}: \B 1\pfun I$ be the distributor given by $(0_{I})_{i}=0$.
%Then $\mathrm{colim}(0_{I}, \mathrm{id}_{I})$ coincides with $\inf I$, via 
%$$
%X(\inf I, x) = \sup_{i\in I}X(i, x)
%$$
%Notice that in general an order-complete $Q$-category needs not be tensored, and thus it needs not be cocomplete.
%
%
%
%
% 
%
%\item \emph{Cauchy-completeness} is the case where $X$ contains the points $\sup \B x, \sup\B y$, for any 
% any two \emph{adjoint} presheaves $\B x,\B y\in [X,Q]$, i.e.~satisfying $0=\inf_{a}\B x_{a}+\B y_{a}$ and $\B x_{a}+ \B x_{b}\geq X(a,b)$. 
% Notice that given a pair $(\B x, \B y)$ we can define a Cauchy sequence by finding $a_{n}$ satisfying 
% $\B x_{a_{n}}+\B y_{a_{n}}\leq \frac{1}{n}$. Indeed this implies $X(a_{n},a_{n+1})\leq \B x_{a_{n}}+\B y_{a_{n+1}}\leq \B x_{a_{n}}+\B y_{a_{n}}+\B x_{a_{n+1}}+\B y_{a_{n+1}}=\frac{1}{n}+\frac{1}{n+1}$.
% Conversely, any (equivalence class of) Cauchy sequences $a_{n}$ yields an adjoint pair given by
% $\B x_{a}= \lim_{n}X(a_{n},a)$ and $\B y_{a}=\lim_{n}X(a,a_{n})$. 
% 
%% 
%% \item \emph{Isbell-completeness} (or \emph{MacNeill cocompleteness}) is the case where $X$ contains the points $\sup \B x, \sup \B y$, for any two presheaves $\B x, \B y\in [X,Q]$ satisfying 
%% $
%% \B x_{a}= \sup_{b\in X}X(b,a)-\B y_{b}
%% $.
%
%
%\end{itemize}
%
%\end{remark}

\subsubsection{Tensors and $Q$-Modules}

Among weighted colimits, one is of big importance for us. 
Any $\epsilon \in Q$ generates a constant distributor $(\epsilon):\B 1\pfun \B 1$, and any point $x\in X$ generates a constant functor $\Delta x: \B 1\to X$.
Given a $Q$-category $X$, a point $x\in X$ and  $\epsilon \in Q$, the \emph{tensor of $x$ and $\epsilon$}, if it exists, is defined as
$$
\epsilon \otimes x := \mathrm{colim}((\epsilon), \Delta x)
$$
A $Q$-category $X$ is \emph{tensored} if for all $x\in X$ and $\epsilon \in Q$, it admits the tensor $\epsilon \otimes x$.

\begin{proposition}
A tensored $Q$-category $X$ is a $Q$-module $(X, \preceq_{X}, \otimes)$.
A cocontinuous functor of cocomplete $Q$-categories is a $Q$-module morphism between the associated $Q$-modules.
\end{proposition}
\begin{proof}
We must show that tensors induce a continuous action. Observe that tensors are characterized by the equation 
\begin{align}\label{eq:tensor}
X(x\otimes \epsilon, x') = X(x,x') -\epsilon
\end{align}
If $\epsilon=0$, then \eqref{eq:tensor} forces $x\otimes \epsilon\simeq x$. 
If $\epsilon=\delta+\eta$, then using the fact that $\alpha-(\epsilon+\delta)=(\alpha-\epsilon)-\delta$ 
we deduce $X((x\otimes \epsilon)\otimes \delta, x')=X(x\otimes \epsilon, x')-\delta=(X(x,x')-\epsilon)-\delta=X(x,x')-(\epsilon+\delta)=X(x\otimes (\epsilon+\delta),x')$, which forces $x\otimes(\epsilon+\delta)\simeq (x\otimes \epsilon)\otimes \delta$.

A cocontinuous functor $f:X\to Y$ commutes with sups and with $\otimes$, and is thus a $Q$-module morphism.
\end{proof}

\begin{lemma}
\begin{itemize}
\item[i.] $\sup_{i\in I}a_{i}-\epsilon= (\sup_{i\in I}a_{i})-\epsilon$.
\item[ii.] $\sup_{i\in I}(a_{i}-\epsilon)-b_{i}= (\sup_{i\in I}a_{i}-b_{i})-\epsilon$.

\end{itemize}
\end{lemma}
\begin{proof}
Let $A= \sup_{i\in I}a_{i}-\epsilon$ and $B= (\sup_{i\in I}a_{i})-\epsilon$.
Let $J\subseteq I$ be the set of indexes $j$ such that $a_{j}>\epsilon$. 
If $J=\emptyset$ then $A=B=0$. Otherwise, 
$A= \sup_{j\in J}a_{j}-\epsilon$ (where ``$-$'' can be interpreted as subtraction on $\BB R$, and 
$B= (\sup_{j\in J}a_{j})-\epsilon$ (again with ``$-$'' being subtraction on $\BB R$), so $A=B$ follows from the continuity of ``$-$'' on $\BB R$.

Let now $A= \sup_{i\in I}(a_{i}-\epsilon)-b_{i}$ and $B= (\sup_{i\in I}a_{i}-b_{i})-\epsilon$.
Let $J\subseteq I$ be the set of indexes $j$ such that $a_{j}> b_{j}+\epsilon$.
If $J=\emptyset$, then $A=0$; suppose $B>0$, then $\sup_{i\in I}a_{i}-b_{i}>\epsilon$, but this implies that we can find $i\in I$ with $a_{i}>b_{i}+\epsilon$, against the assumption, so also $B=0$ holds. If $J$ is non-empty, then 
$A= \sup_{j\in J}(a_{j}-\epsilon)-b_{j}$, where ``$-$'' is not subtraction on $\BB R$ and 
$B= (\sup_{j\in J}a_{j}-b_{j})-\epsilon$, again with ``$-$'' usual subtraction, so $A=B$ follows from the continuity of ``$-$'' on $\BB R$.
 \end{proof}

\begin{lemma}
In any cocomplete $Q$-category, $x\otimes \epsilon \simeq \sup(  \C Y(x)+\epsilon  )$.
In the cocomplete $Q$-category $[X,Q]$, $\B x\otimes \epsilon= \B x+\epsilon$.
\end{lemma}
\begin{proof}
We have 
\begin{align*}
X(\sup(\C Y(x)+\epsilon), x')& =\sup_{y\in X}X(z,x')- (\C Y(x)(z)+\epsilon)\\
&= \sup_{y\in X}X(z,x')-(X(z,x)+\epsilon)\\
&= (\sup_{y\in X}X(z,x')-X(z,x))-\epsilon\\
&= X(x,x')-\epsilon
\end{align*}
which shows $x\otimes \epsilon=\sup(\C Y(x)+\epsilon)$. In $[X,Q]$ we have 
$[X,Q](\B x+\epsilon, \B x')=\sup_{a\in X}(\B x_{a}+\epsilon)-\B x'_{a}= (\sup_{a\in X}\B x_{a}-\B x'_{a})-\epsilon= [X,Q](\B x, \B x')-\epsilon$, which shows $\B x\otimes \epsilon \simeq \B x+\epsilon$, and since $[X,Q]$ is skeletal, $\B x\otimes \epsilon=\B x+\epsilon$.
\end{proof}

The dual notion of tensors is the \emph{cotensor} $x\multimapinv \epsilon$. Formally, it is defined as a \emph{weighted limit} (whose definition is dual to that of weighted colimit but we do not give details here), and characterized by the equation
\begin{align*}
X(x', x\multimapinv \epsilon)= X(x',x)-\epsilon
\end{align*}
In other words, in a tensored and cotensored $Q$-category we have $X(x\otimes \epsilon,y)= X(x,y\multimapinv \epsilon)$.

\begin{example}
The $Q$-category $[X,Q]$ is cotensored, with $\B x\multimapinv \epsilon:= \B x-\epsilon$. Indeed we have $[X,Q](\B x, \B y\multimapinv \epsilon)=\sup_{a\in X}(\B y_{a}-\epsilon)-\B x_{a}=(\sup_{a\in X}\B y_{a}-\B x_{a})-\epsilon= [X,Q](\B x, \B y)-\epsilon$.
\end{example}

\begin{definition}
A $Q$-category $X$ is \emph{order-complete} if it is a sup-lattice with respect to the order $\preceq_{X}$ (i.e.~all joins exist).
\end{definition}

\begin{lemma}\label{lemma:supinf}
Let $X$ be a $Q$-category. If $X$ is order-complete, then 
\begin{itemize}
\item if $X$ is co-tensored, $X(\bigvee_{i}x_{i},y)=  \sup_{i}X(x_{i},y) $;
\item if $X$ is tensored, $
X(x,\bigvee_{i}y_{i})=  \inf_{i}X(x,y_{i})$.

\end{itemize}
\end{lemma}
\begin{proof}
We only prove the second claim, the first being proved similarly.

 Let us show that $z\preceq_{X}z'$ iff for all $w\in X$, $X(w,z')\leq X(w,z)$: 
 on one direction we have $X(w,z')\leq X(w,z)+X(z,z')=X(w,z)$; on the other direction, 
 we have $X(z,z')\leq X(z,z)=0$.

 Using this, since $y_{i}\preceq_{X}y:=\bigvee_{i}y_{i}$ we deduce 
 $X(x,y_{i})\leq X(x,y)$, and thus $X(x,y)\geq \inf_{i}X(x,y_{i})$.

 For the converse direction, we argue as follows: let $X(x,y_{i})\leq \epsilon$ hold for all $i\in I$; then $0=X(x,y_{i})-\epsilon= X(x\otimes\epsilon,y_{i})$. Thus $x\otimes\epsilon\preceq_{X}y_{i}$, and thus
 $x\otimes\epsilon\preceq_{X}y$, that is $X(x\otimes\epsilon,y)=X(x,y)-\epsilon=0$, and consequently $X(x,y)\leq \epsilon$. 
 By letting $\epsilon:=X(x,y_{i})$ we conclude then $X(x,y)\leq X(x,y_{i})$, and thus $X(x,y)\leq \inf_{i}X(x,y_{i})$.

%
%First, if $
%By definition $x:=\bigvee_{i}x_{i}$ is characterized by (1) $X(x_{i},x)=0$ for all $i\in I$, and 
% (2) $X(x,y)=0$, for all $y$ such that $X(x_{i},y)=0$ holds for all $i\in I$.
%% 
%% Let us show that $z\preceq_{X}z'$ implies $X(z',y) \leq X(z,y)$: 
%% from $z\preceq_{X}z'$ we deduce $X(z,z')=0$, whence $
%% 
% Let now $y\in X$; 
% then $X(x_{i},y)\leq X(x_{i},x)+X(x,y)=X(x,y)$, which implies $\sup_{i}X(x_{i},y)\leq X(x,y)$.
% 
%
% 
% 
% Suppose now that, for some $i\in I$, $X(x_{i},y)>0$; then $X(x,y)\leq X(x,x_{i})+X(x_{i},y)$
% 
% 
%If $X(x_{i},y) \leq \epsilon$, for all $i\in I$, then 
%$0= X(x_{i},y)-\epsilon = X(x_{i},y\multimapinv \epsilon)$. Thus 
%$x_{i}\preceq_{X}y\multimapinv\epsilon$ and we deduce 
%$x\preceq_{X}y\multimapinv \epsilon$. Hence
%$0= X(x, y\multimapinv \epsilon)=X(x,y)-\epsilon$ and consequently
%$ X(x,y)\leq \epsilon$.

\end{proof}

\begin{proposition}\label{prop:tencoten}
If a $Q$-category $X$ is tensored, cotensored and order-complete, then it is cocomplete.
\end{proposition}
\begin{proof}
For all $\B x\in [X,Q]$, let $\sup \B x:= \bigvee_{a\in X}a\otimes\B x_{a}$. 
Let us check that $X(\sup \B x, b)= \sup_{a\in X}X(a,b)-\B x_{a}$, using Lemma \ref{lemma:supinf}:
\begin{align*}
X(\sup \B x, b) &= \sup_{a\in X}X(a\otimes \B x_{a},b)\\
&=\sup_{a\in X}X(a,b)-\B x_{a}
\end{align*}
We can thus conclude using Proposition \ref{prop:yonedasup}.
\end{proof}

\begin{proposition}\label{prop:tenfun}
Let $X,Y$ be two tensored $Q$-categories, and $f:X\to Y$ be a function.
\begin{itemize}
\item[i.] $f$ is a functor iff $f$ is order-preserving and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon \preceq_{Y} f(x\otimes \epsilon)$.

\item[ii.] $f$ is a cocontinuous functor iff $f$ commutes with joins and for all $x\in X$ and $\epsilon\in Q$, $f(x)\otimes \epsilon = f(x\otimes \epsilon)$.
\end{itemize}
\end{proposition}
\begin{proof}
\begin{itemize}
\item[i.] If $f$ is a functor then 
\begin{align*}
Y(f(x)\otimes \epsilon, f(x\otimes \epsilon))&= Y(f(x), f(x\otimes \epsilon)) -\epsilon \\
&\leq X(x, x\otimes \epsilon)-\epsilon \\
&= X(x\otimes \epsilon, x\otimes \epsilon)=0
\end{align*}
so $Y(f(x)\otimes \epsilon, f(x\otimes \epsilon))=0$, which implies
$f(x)\otimes \epsilon \preceq_{X}f(x\otimes \epsilon)$. 
Moreover, if $x\preceq_{X}x'$, then $0\geq X(x,x')\geq Y(f(x),f(x'))$, whence 
$f(x)\preceq_{Y}f(x')$, so $f$ is order-preserving.

Conversely, for all $x,x'\in X$, 
\begin{align*}
X(x\otimes X(x,x'), x') &=X(x,x')-X(x,x')=0 
\end{align*}
thus $x\otimes X(x,x') \preceq_{X}x'$. Since $f$ is order-preserving, it follows that
\begin{align*}
f(x)\otimes X(x,x') \preceq_{Y}f(x\otimes X(x,x'))\preceq_{Y}f(x') 
\end{align*}
which implies that 
\begin{align*}
Y(f(x),f(x')) - X(x,x') = Y(f(x)\otimes X(x,x'), f(x'))=0
\end{align*}
that is $Y(f(x),f(x'))\leq X(x,x')$, so $f$ is a functor.

>>>>

\subsubsection{The language $\BSTLC$}
TO BE ADDED

\subsubsection{$(\LREL, !_{n})$ is a model of Bounded Linear Logic}

Given SMCs $\C C,\C D$,  let $\B{SMC}_{l}(\C C, \C D)$ indicate the category of symmetric lax monoidal functors and monoidal natural transformations between them.
$\B{SMC}_{l}(\C C, \C D)$  is itself a SMC, with monoidal structure defined pointwise.
